Sex-Specific Evolution of the Meiotic Recombination Rate

### 2020-06-22 doc/MS_v3.Rmd
some more updates to intro and edits

@@

  **Due to it's role in facilitating chromosomes disjunction, the genome-wide recombination rate can be thought of as a cell-based metric which is tightly connected to organisms fitness and fertility.**  (thus clear target of evolution). Ensuring at least one crossover per chrm while minimizing the rates of DNA damage and ectopic recombination are thought to regulate the lower and upper thresholds for a genome-wide recombiantion rate [@nagaoka2012] (Inoue). Yet, within these thresholds the genome-wide recombination rate still varies across species and individuals. Sex is one of the most notable axes of this form of variation. 
  There are two classes of sexual dimorphism in recombiantion rates; the first is called achiasmy and refered to where one sex recombines (Bell 1982 – Huxley 1928, Lenormand Duthiel 2005). The second, which is the focus of this manuscript, is the instance where both sexes recombine (Lenormand and Duthiel 2003). An understanding of how sex shapes the evolution of recombination cannot be achieved with available data. Comprehensive comparisons of female and male recombination rates usually come from outbred populations humans [@Kong2004;, @Kong2008; @Kong2014; @halldorsson2019], dog [@wong2010dog],  cattle [@ma2015_cattle; @Shen2018_cattle], (other sheep?) sheep [@johnston2016_soay], and mouse [@srivastava2017CC] in which the role of sex is confounded with the contributions of genetic variation.

@@

  The genome-wide recombiantion rate can be decomposed into single chromosome patterns, refered to as recombination landscapes. **(at this scale –there are also relevant features of heterochiasmy).** Recently two specific features have been highlighted as conserved patterns related heterochiasmy, specifically the recombination landscape. In a review of data from mouse, celegans and worms with meiotic chromosome morphilogy (axis-loop structure) quantified for both sexes, @CahoonLibuda2019 find that the sex with longer chromosome axis generally has higher recombination rates. A survey of 51 species shows there are conserved broad scale recombination landscape differences. Generally males have telomere-bias crossover placement and females have more uniform placement [@sardell_sex_2020]. These observations raise the question of how conservation in sexually dimorphic patterns are maintained with the gwRR and it’s decomposed traits.

@@

  In our comparisons of recombiantion landscapes across sex, our results confirm previously reported patterns for sex differences in the general recombination landscape [@sardell_sex_2020]. Intriguingly,  these patterns hold even in the case of genome more crossovers, but general landscape patterns still maintain sex-specific patterns.

@@

<new 1/n>

@@

<new 2/n>
To measure bivalent SC length, two image analysis algorithms were used. The first algorithm estimates the total (summed) SC length across bivalents for individual cells (@wang2019_SC). The second algorithm estimates the SC length of individual bivalents (@peterson2019). Both algorithms apply a ‘skeletonizing’ transformation to synapsed chromosomes that produces a single, pixel-wide ‘trace’ of the bivalent shape. Total SC length per cell was quantified from pachytene cell images (@wang2019_SC).

@@

<new 3/n>
To account for confounding effects of sex chromosomes from pooled samples of bivalents, we also considered a reduced data set including only bivalents with SC lengths below the 2nd quartile in cells with at least 17 of 20 single bivalent measures. This “short bivalent” data set included the four or five shortest bivalents and excluded the X bivalent in oocytes. A total of `r sum(Short.biv_mouse_table$n_obs )` short bivalents were isolated from `r sum(  (Short.biv_mouse_table  %>% filter(sex=="female") )$Ncells)` oocytes and `r sum(  (Short.biv_mouse_table  %>% filter(sex=="male") )$Ncells)` spermatocytes. Although this smaller data set has decreased power, it offers a more comparable set of single bivalents to compare between the sexes. A “long bivalent” data set was formed from those bivalents above the 4th quartile in SC lengths per cell. A total of `r sum(Long.biv_mouse_table$n_obs)` long bivalents were isolated from `r sum(  (Long.biv_mouse_table  %>% filter(sex=="female") )$Ncells)` oocytes and `r sum(  (Long.biv_mouse_table  %>% filter(sex=="male") )$Ncells)` spermatocytes.

<new 4/n>
To examine crossover interference, the distance (in SC units) between MLH1 foci (inter-focal distance; IFD~raw~) was measured for those single bivalents containing two MLH1 foci. A normalized measure of interference (IFD~norm~) was computed by dividing IFD~raw~ by SC length on a per-bivalent basis.

<new 5/n>
We used a series of statistical models to interpret patterns of variation in the recombination traits we measured (Table 2). We used mouse average as the dependent variable in all analyses. We first constructed a linear mixed model (M1) using lmer() from the lmer4 package [@lme4] in R (v3.5.2) [@Rstudio]. In this model, strain was coded as a random effect, with significance evaluated using a likelihood ratio test (using exactRLRT() from RLRsim [@RLRsim]). Subspecies, sex, and their interaction were coded as fixed effects, with significance evaluated using a chi-square test comparing the full and reduced models (drop1() and anova()) [@lme4]. As each observation was at the level of a single mouse (mouse average), it was uniquely coded within the dataset, nesting was implicit, each strain only occurs within one strain, and was not explicitly coded in our linear and mixed models. We used the subspecies effect to quantify divergence between subspecies and the (random) strain effect to quantify variation within subspecies in a sex-specific manner. In separate analyses using model M1, we considered mouse averages as dependent variables for each of the following traits: MLH1 count per cell, total SC length per cell, single bivalent SC length per cell, IFD~raw~, IFD~norm~, and average MLH1 position (for single-focus bivalents). Four additional linear models containing only fixed effects (M2-M5) (Table 2) were used to further investigate results obtained from model M1.

@@

The effect of subspecies is no longer significant in a model treating all factors as fixed effects **(M2; Musc p = `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][2]`, Mol p =  `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][3]`)**, highlighting strain and sex as salient variables. Two strains exhibit particularly strong effects on MLH1 focus count (M3; *domesticus^G^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][3]` ; *domesticus^LEW^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][4]`), with sex-strain interactions involving three strains (M3; *domesticus^G^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][10]`; *molossinus^MSM^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][13]` ; *musculus^PWD^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][12]`).

In separate analyses of males (M4; n = 80 - `r length(male_mouse.avs_for.analysis$mouse)`), three strains disproportionately shape MLH1 focus count (as observed in Figure 1C): *musculus^PWD^* (p = `r sum_MLH1.male.glm.M4$coefficients[,4][7]` ; effect = `r sum_MLH1.male.glm.M4$coefficients[7]` ), *molossinus^MSM^* (p = `r sum_MLH1.male.glm.M4$coefficients[,4][8]`; effect = `r sum_MLH1.male.glm.M4$coefficients[8]`), and *musculus^SKIVE^* (p = `r sum_MLH1.male.glm.M4$coefficients[,4][9]`; effect = `r sum_MLH1.male.glm.M4$coefficients[9]` ). These three strains point to substantial evolution in the genome-wide recombination rate in spermatocytes; we subsequently refer to them as “high-recombination” strains.

In females (M4; n= 76 - `r length(female_mouse.avs_for.analysis$mouse)`), four strains affect MLH1 focus count: *domesticus^G^* (p = `r sum_MLH1.female.spc.glm.M5$coefficients[,4][2]`; effect = `r sum_MLH1.female.spc.glm.M5$coefficients[2]`), *molossinus^MSM^* (p = `r sum_MLH1.female.spc.glm.M5$coefficients[,4][5]`; effect = `r sum_MLH1.female.spc.glm.M5$coefficients[5]`), *domesticus^LEW^* (p = `r sum_MLH1.female.spc.glm.M5$coefficients[,4][3]` ; effect = `r sum_MLH1.female.spc.glm.M5$coefficients[3]`), and *musculus^PWD^* (p = `r sum_MLH1.female.spc.glm.M5$coefficients[,4][4]`; effect = `r sum_MLH1.female.spc.glm.M5$coefficients[4]`). Strain effect sizes in females are modest in magnitude compared to those in males.

Together, these results demonstrate that the genome-wide recombination rate evolves in a highly sex-specific manner.

>>>>

Meiosis can be reduced to the expression (2n -> 4n -> 2n -> 1n) which reflects the process of a diploid genome producing haploid gametes. The total number of COs at this stage of meiosis is equal to the genome-wide recombination rate (gwRR). Meiosis, and the mixing of haploid gametes, is an important process for evolution, because it can determine a populations’ response to selection by transferring beneficial mutations onto novel genetic backgrounds or by breaking linkage of negative mutations from beneficial genetic backgrounds [@charlesworth1993]. This process shapes the genomic patterns of genetic variation, with high recombination areas of genome having more nucleotide variation while areas of low recombination have decreased genetic variation [@begun1992; @nachman2012].

@@

  **Due to it's role in facilitating chromosomes disjunction, the genome-wide recombination rate can be thought of as a cell-based metric which is tightly connected to organisms fitness and fertility.**  (thus clear target of evolution). Ensuring at least one crossover per chrm while minimizing the rates of DNA damage and ectopic recombination are thought to regulate the lower and upper thresholds for a genome-wide recombination rate [@nagaoka2012] (Inoue). Yet, within these thresholds the genome-wide recombination rate still varies across species and individuals. Sex is one of the most notable axes of this form of variation. 
  There are two classes of sexual dimorphism in recombination rates; the first is called achiasmy and refered to where one sex recombines (Bell 1982 – Huxley 1928, Lenormand Duthiel 2005). The second, which is the focus of this manuscript, is the instance where both sexes recombine (Lenormand and Duthiel 2003). An understanding of how sex shapes the evolution of recombination cannot be achieved with available data. Comprehensive comparisons of female and male recombination rates usually come from outbred populations humans [@Kong2004;, @Kong2008; @Kong2014; @halldorsson2019], dog [@wong2010dog],  cattle [@ma2015_cattle; @Shen2018_cattle], (other sheep?) sheep [@johnston2016_soay], and mouse [@srivastava2017CC] in which the role of sex is confounded with the contributions of genetic variation.

@@

  The genome-wide recombination rate can be decomposed into single chromosome patterns, refered to as recombination landscapes. **(at this scale –there are also relevant features of heterochiasmy).** Recently two specific features have been highlighted as conserved patterns related heterochiasmy, specifically the recombination landscape. In a review of data from mouse, celegans and worms with meiotic chromosome morphilogy (axis-loop structure) quantified for both sexes, @CahoonLibuda2019 find that the sex with longer chromosome axis generally has higher recombination rates. A survey of 51 species shows there are conserved broad scale recombination landscape differences. Generally males have telomere-bias crossover placement and females have more uniform placement [@sardell_sex_2020]. These observations raise the question of how conservation in sexually dimorphic patterns are maintained with the gwRR and it’s decomposed traits.

@@

  In our comparisons of recombination landscapes across sex, our results confirm previously reported patterns for sex differences in the general recombination landscape [@sardell_sex_2020]. Intriguingly,  these patterns hold even in the case of genome more crossovers, but general landscape patterns still maintain sex-specific patterns.

@@

@@

To measure bivalent SC length, two image analysis algorithms were used. The first algorithm estimates the total (summed) SC length across bivalents for individual cells (@wang2019_SC). The second algorithm estimates the SC length of individual bivalents (@peterson2019). Both algorithms apply a ‘skeletonizing’ transformation to synapsed chromosomes that produces a single, pixel-wide ‘trace’ of the bivalent shape. Total SC length per cell was quantified from pachytene cell images (@wang2019_SC).

@@

To account for confounding effects of sex chromosomes from pooled samples of bivalents, we also considered a reduced data set including only bivalents with SC lengths below the 2nd quartile in cells with at least 17 of 20 single bivalent measures. This “short bivalent” data set included the four or five shortest bivalents and excluded the X bivalent in oocytes. A total of `r sum(Short.biv_mouse_table$n_obs )` short bivalents were isolated from `r sum(  (Short.biv_mouse_table  %>% filter(sex=="female") )$Ncells)` oocytes and `r sum(  (Short.biv_mouse_table  %>% filter(sex=="male") )$Ncells)` spermatocytes. Although this smaller data set has decreased power, it offers a more comparable set of single bivalents to compare between the sexes. A “long bivalent” data set was formed from those bivalents above the 4th quartile in SC lengths per cell. A total of `r sum(Long.biv_mouse_table$n_obs)` long bivalents were isolated from `r sum(  (Long.biv_mouse_table  %>% filter(sex=="female") )$Ncells)` oocytes and `r sum(  (Long.biv_mouse_table  %>% filter(sex=="male") )$Ncells)` spermatocytes.

To examine crossover interference, the distance (in SC units) between MLH1 foci (inter-focal distance; IFD~raw~) was measured for those single bivalents containing two MLH1 foci. A normalized measure of interference (IFD~norm~) was computed by dividing IFD~raw~ by SC length on a per-bivalent basis.

We used a series of statistical models to interpret patterns of variation in the recombination traits we measured (Table 2). We used mouse average as the dependent variable in all analyses. We first constructed a linear mixed model (M1) using lmer() from the lmer4 package [@lme4] in R (v3.5.2) [@Rstudio]. In this model, strain was coded as a random effect, with significance evaluated using a likelihood ratio test (using exactRLRT() from RLRsim [@RLRsim]). Subspecies, sex, and their interaction were coded as fixed effects, with significance evaluated using a chi-square test comparing the full and reduced models (drop1() and anova()) [@lme4]. As each observation was at the level of a single mouse (mouse average), it was uniquely coded within the dataset, nesting was implicit, each strain only occurs within one strain, and was not explicitly coded in our linear and mixed models. We used the subspecies effect to quantify divergence between subspecies and the (random) strain effect to quantify variation within subspecies in a sex-specific manner. In separate analyses using model M1, we considered mouse averages as dependent variables for each of the following traits: MLH1 count per cell, total SC length per cell, single bivalent SC length per cell, IFD~raw~, IFD~norm~, and average MLH1 position (for single-focus bivalents). Four additional linear models containing only fixed effects (M2-M5) (Table 2) were used to further investigate results obtained from model M1.

@@

The effect of subspecies is no longer significant in a model treating all factors as fixed effects (M2; Musc p = `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][2]`, Mol p =  `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][3]`), highlighting strain and sex as salient variables. Two strains exhibit particularly strong effects on MLH1 focus count (M3; *domesticus^G^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][3]` ; *domesticus^LEW^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][4]`), with sex-strain interactions involving three strains (M3; *domesticus^G^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][10]`; *molossinus^MSM^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][13]` ; *musculus^PWD^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][12]`).

In separate analyses of males (M4; n = `r length(male_mouse.avs_for.analysis$mouse)`), three strains disproportionately shape MLH1 focus count (as observed in Figure 1C): *musculus^PWD^* (p = `r sum_MLH1.male.glm.M4$coefficients[,4][7]` ; effect = `r sum_MLH1.male.glm.M4$coefficients[7]` ), *molossinus^MSM^* (p = `r sum_MLH1.male.glm.M4$coefficients[,4][8]`; effect = `r sum_MLH1.male.glm.M4$coefficients[8]`), and *musculus^SKIVE^* (p = `r sum_MLH1.male.glm.M4$coefficients[,4][9]`; effect = `r sum_MLH1.male.glm.M4$coefficients[9]` ). These three strains point to substantial evolution in the genome-wide recombination rate in spermatocytes; we subsequently refer to them as “high-recombination” strains. In females (M4; n= `r length(female_mouse.avs_for.analysis$mouse)`), four strains affect MLH1 focus count: *domesticus^G^* (p = `r sum_MLH1.female.spc.glm.M5$coefficients[,4][2]`; effect = `r sum_MLH1.female.spc.glm.M5$coefficients[2]`), *molossinus^MSM^* (p = `r sum_MLH1.female.spc.glm.M5$coefficients[,4][5]`; effect = `r sum_MLH1.female.spc.glm.M5$coefficients[5]`), *domesticus^LEW^* (p = `r sum_MLH1.female.spc.glm.M5$coefficients[,4][3]` ; effect = `r sum_MLH1.female.spc.glm.M5$coefficients[3]`), and *musculus^PWD^* (p = `r sum_MLH1.female.spc.glm.M5$coefficients[,4][4]`; effect = `r sum_MLH1.female.spc.glm.M5$coefficients[4]`). Strain effect sizes in females are modest in magnitude compared to those in males. Together, these results demonstrate that the genome-wide recombination rate evolves in a highly sex-specific manner.
